---
title: "Class 05: Data Visualization with GGPLOT"
author: "Diego Diaz (PID: A17328629)"
format: pdf
toc: TRUE
---

## Background

There are lots of ways to make figures in R. These include so-called "base R" graphics (e.g. `plot()`) and tons of add-on packages like **ggplot2**. 

For example here we make the same plot with both:

```{r}
head(cars)
```
```{r}
plot(cars)
```

First I need to install the package with the command `install.packages()`.

> **N.B.** We never run an install in a quarto code chunck or we will end up re-installing packages many, many times - which is not what we want!

Everytime we want to use of these "add-on" packages we need to load it up in R with the `library()` function:

```{r}
library(ggplot2)
```


```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things:

- The **data**, the stuff you want plotted.
- The **aes**thetics, how data is mapped to the plot.
- The **geom**etry, the type of plot.

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() 
```

```{r}
p <- ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE) +
  labs(title="Stopping Distance of Old Cars",
       subtitle = "Data from the `cars` object", 
       x = "Speed (mph)",
       y = "Distance (ft)")
```
render it out

```{r}
p + theme_bw()
```

## Gene Expression Plot

We can read the input data from the class website.

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot:

```{r}
ggplot(genes) +
       aes(Condition1, Condition2) +
       geom_point()
```

```{r}
table(genes$State)
```

Version 2: Let's color by `State`, so we can compare the "up" and "down" significant genes compared to all the "unchanging" genes.

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point()
```

Version 3: Let's modify the default colors to something we like.

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point() +
  scale_color_manual(values= c("#008080", "#636363", "purple")) +
  labs(x="Control (no drug)",
       y="Drug",
       title="Gene Expression Changes upon GLP-1 drug")
```

## Going Further

Let's have a look at the famous **gapminder** data set.

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=(0.35))
```

Let's "facet" (i.e. make a separate plot) by continents rather than the big hot mess above.

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=(0.35)) +
  facet_wrap(~continent)
```

## Custom Plots

How big is this gapminder dataset?

```{r}
nrow(gapminder)
```
I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me.

First I need to install it and load it up.
`install.packages("dplyr")` and:
```{r}
library(dplyr)

gapminder2007 <- filter(gapminder, year == 2007)
head(gapminder2007)
```


```{r}
filter(gapminder2007, country == "Ireland")
```

```{r}
filter(gapminder, year == 1977, country == "United States")
```

> Q. Make a plot comparing 1977 and 2007 for all countries

```{r}
gapminderplot <- filter(gapminder, year %in% c(1977, 2007)) 
ggplot(gapminderplot) +
  aes(gdpPercap, lifeExp, col = continent) +
  geom_point() +
  facet_wrap(~year) +
  labs(x="GDP per Capita",
       y="Life Expectancy")
```

